Well-quasi-orderings on word languages

### 2025-10-15 src/appendix.tex
Move some stuff to appendix

>>>>

\begin{proofof}{small-ordinal-invariants:lem}
	Let $N_0$ be a bound such that $w_{\geq N_0}$ is \kl{uniformly recurrent}.
	Let us write $\infset{w}$ the set of finite infixes of $w$.
	We prove that $\oWidth{\infset{w}} \leq \omegaOrd + N_0$.
	Let $u_1 \infleq w$ be a finite word.

	If $u_1$ is an \kl{infix} of $w_{\geq N_0}$, then there exists $k \geq 1$
	such that $u_1$ is an \kl{infix} of every word of size at least $k$. In
	particular, there is finite bound on the length of every sequence of
	incomparable elements starting with $u_1$. We conclude in particular that
	$\infset{w} \setminus \upset{u_1}$ has a finite \kl{ordinal width}.

	Otherwise, $u_1$ can only be found \emph{before} $N_0$. In this case, we
	consider a second element of a \kl{bad sequence} $u_2 \infleq w$, which is
	incomparable with $u_1$ for the \kl{infix relation}. If $u_2$ is an
	\kl{infix} of $w_{\geq N_0}$, then we can conclude as before. Otherwise,
	notice that $u_1$ and $u_2$ cannot start at the same position in $w$
	(because they are incomparable). Continuing this argument, we conclude that
	there are at most $N_0$ elements starting before $N_0$
	at the start of any sequence of
	incomparable elements in $\infset{w}$. We conclude that
	$\oWidth{\infset{w}} \leq \omegaOrd + N_0$.

	Let us now justify that this bound is tight.
	The \kl{Thue-Morse sequence} over a binary
	alphabet $\set{a,b}$ has \kl{ordinal width} $\omegaOrd$
	from \cref{thue-morse-ordinal:lemma}.
	Given a number $N_0
	\in \Nat$, one can construct an arbitrarily long \kl{antichain} of words for
	the \kl{infix relation} by using a new letter $c$. When concatenating this
	(finite) antichain as a prefix of the \kl{Thue-Morse sequence}, one obtains a
	new (infinite) word $w$. It is clear that the \kl{ordinal width} of
	$\infset{w}$ is now at least $\omegaOrd + N_0$.
\end{proofof}

\begin{proofof}{from-bi-to-single:lem}
	Given a bi-infinite word $w \in \Sigma^{\Rel}$, recall that we can consider $w_+ \in
	\Sigma^\Nat$ and $w_- \in \Sigma^\Nat$ the two infinite words obtained as
	follows: for all $i \in \Nat$, $(w_+)_i = w(i)$ and $(w_-)_i = w(-i)$. Note
	that the two share the letter at position $0$.

	To obtain the upper bound of $3 \cdot \omegaOrd$, we can consider the same
	argument as for \cref{small-ordinal-invariants:lem}. We let $N_0$ be such
	that $w_{\geq N_0}$ and $(w_-)_{\geq N_0}$ are \kl{uniformly recurrent}
	words. In any sequence of incomparable elements of $\infset{w}$, there are
	less than $N_0^2$ elements that are found in $(w_{< N_0})_{> -N_0}$. Then,
	one has to pick a finite \kl{infix} in either $w_{\geq N_0}$ or $w_{\leq
		-N_0}$. Because of \cref{small-ordinal-invariants:lem}, any sequence of
	incomparable elements of these two infinite words has length bounded based
	on the choice of the first element of that sequence. This means that the
	\kl{ordinal width} of $\infset{w}$ is at most $\omegaOrd + \omegaOrd +
	N_0^2$. We conclude that $\oWidth{\infset{w}} < 3 \cdot \omegaOrd$.

	Let us briefly argue that the bound is tight. Indeed, one can
	construct a bi-infinite word $w$ by concatenating a reversed \kl{Thue-Morse
		sequence} on a binary alphabet $\set{a,b}$, a finite antichain of arbitrarily
	large size over a distinct alphabet $\set{c,d}$, and then a \kl{Thue-Morse
		sequence} on a binary alphabet $\set{e,f}$. The \kl{ordinal width} of the set
	of \kl{infixes} of $w$ is then at least $2 \cdot \omegaOrd + K$, where $K$ is the
	size of the chosen antichain, following the same argument as in the proof of
	\cref{small-ordinal-invariants:lem}, using \cref{thue-morse-ordinal:lemma}.
\end{proofof}

### 2025-10-15 src/infixes.tex
Move some stuff to appendix

We are now ready to conclude the proof of \cref{dw-closed-wqo-charac:thm}.
\begin{proofof}{dw-closed-wqo-charac:thm}
  It is clear that the set of finite \kl{infixes} of a finite, infinite or
  bi-infinite \kl{ultimately uniformly recurrent} word is \kl{well-quasi-ordered}
  for the \kl{infix relation} thanks to \cref{ultimately-uniformly-recurrent:lem}.

@@

Let us now analyse the \kl{ordinal invariants} of the set of finite
\kl{infixes} of an \kl{ultimately uniformly recurrent} infinite word, in order
to derive the bounds of \cref{small-ordinal-invariants:thm}.

\begin{lemma}
    \label{small-ordinal-invariants:lem}
    Let $w \in \Sigma^\Nat$ be an \kl{ultimately uniformly recurrent} word.
    Then, the set of finite infixes of $w$ has \kl{ordinal width}
    less than $2 \cdot \omegaOrd$.
    Furthermore, this bound is tight. 
\end{lemma}
\begin{proof}
    Let $N_0$ be a bound such that $w_{\geq N_0}$ is \kl{uniformly recurrent}.
    Let us write $\infset{w}$ the set of finite infixes of $w$.
    We prove that $\oWidth{\infset{w}} \leq \omegaOrd + N_0$.
    Let $u_1 \infleq w$ be a finite word.

    If $u_1$ is an \kl{infix} of $w_{\geq N_0}$, then there exists $k \geq 1$
    such that $u_1$ is an \kl{infix} of every word of size at least $k$. In
    particular, there is finite bound on the length of every sequence of
    incomparable elements starting with $u_1$. We conclude in particular that
    $\infset{w} \setminus \upset{u_1}$ has a finite \kl{ordinal width}.

    Otherwise, $u_1$ can only be found \emph{before} $N_0$. In this case, we
    consider a second element of a \kl{bad sequence} $u_2 \infleq w$, which is
    incomparable with $u_1$ for the \kl{infix relation}. If $u_2$ is an
    \kl{infix} of $w_{\geq N_0}$, then we can conclude as before. Otherwise,
    notice that $u_1$ and $u_2$ cannot start at the same position in $w$
    (because they are incomparable). Continuing this argument, we conclude that
    there are at most $N_0$ elements starting before $N_0$
    at the start of any sequence of
    incomparable elements in $\infset{w}$. We conclude that
    $\oWidth{\infset{w}} \leq \omegaOrd + N_0$.

  Let us now justify that this bound is tight.
  The \kl{Thue-Morse sequence} over a binary
  alphabet $\set{a,b}$ has \kl{ordinal width} $\omegaOrd$
  from \cref{thue-morse-ordinal:lemma}.
    Given a number $N_0
  \in \Nat$, one can construct an arbitrarily long \kl{antichain} of words for
  the \kl{infix relation} by using a new letter $c$. When concatenating this
  (finite) antichain as a prefix of the \kl{Thue-Morse sequence}, one obtains a
  new (infinite) word $w$. It is clear that the \kl{ordinal width} of
  $\infset{w}$ is now at least $\omegaOrd + N_0$.
\end{proof}

\begin{lemma}
    \label{from-bi-to-single:lem}
    Let $w \in \Sigma^\Rel$ be a bi-infinite word. Then, the set of finite
    infixes of $w$ is \kl{well-quasi-ordered} for the \kl{infix relation} if
    and only if $w_+$ and $w_-$ are two \kl{ultimately uniformly recurrent}
    words. In this case, the \kl{ordinal width}
    of the set of finite infixes of $w$ is less than $3 \cdot \omegaOrd$,
    and this bound is tight.
\end{lemma}
\begin{proof}
  Given a bi-infinite word $w \in \Sigma^{\Rel}$, recall that we can consider $w_+ \in
\Sigma^\Nat$ and $w_- \in \Sigma^\Nat$ the two infinite words obtained as
follows: for all $i \in \Nat$, $(w_+)_i = w(i)$ and $(w_-)_i = w(-i)$. Note
that the two share the letter at position $0$.

    To obtain the upper bound of $3 \cdot \omegaOrd$, we can consider the same
    argument as for \cref{small-ordinal-invariants:lem}. We let $N_0$ be such
    that $w_{\geq N_0}$ and $(w_-)_{\geq N_0}$ are \kl{uniformly recurrent}
    words. In any sequence of incomparable elements of $\infset{w}$, there are
    less than $N_0^2$ elements that are found in $(w_{< N_0})_{> -N_0}$. Then,
    one has to pick a finite \kl{infix} in either $w_{\geq N_0}$ or $w_{\leq
    -N_0}$. Because of \cref{small-ordinal-invariants:lem}, any sequence of
    incomparable elements of these two infinite words has length bounded based
    on the choice of the first element of that sequence. This means that the
    \kl{ordinal width} of $\infset{w}$ is at most $\omegaOrd + \omegaOrd +
    N_0^2$. We conclude that $\oWidth{\infset{w}} < 3 \cdot \omegaOrd$.

  Let us briefly argue that the bound is tight. Indeed, one can
  construct a bi-infinite word $w$ by concatenating a reversed \kl{Thue-Morse
  sequence} on a binary alphabet $\set{a,b}$, a finite antichain of arbitrarily
  large size over a distinct alphabet $\set{c,d}$, and then a \kl{Thue-Morse
  sequence} on a binary alphabet $\set{e,f}$. The \kl{ordinal width} of the set
  of \kl{infixes} of $w$ is then at least $2 \cdot \omegaOrd + K$, where $K$ is the
  size of the chosen antichain, following the same argument as in the proof of
  \cref{small-ordinal-invariants:lem}, using \cref{thue-morse-ordinal:lemma}.
\end{proof}

We are now ready to conclude the proof of \cref{small-ordinal-invariants:thm}.
\begin{proofof}{small-ordinal-invariants:thm}
    It is always true that the \kl{ordinal height} of a language over a finite
    alphabet is at most $\omegaOrd$. Let us now consider a
    \kl{well-quasi-ordered} language $L$ that is \kl{downwards closed} for the
    \kl{infix relation}. Applying 
    \cref{dw-closed-wqo-charac:thm}, we can write $L = \bigcup_{i = 1}^n L_i$ where
    each $L_i$ is the set of finite \kl{infixes} of a finite, infinite or
    bi-infinite \kl{ultimately uniformly recurrent} word $w_i$.
   We can then
    directly conclude that $\oWidth{L_i}$ less than $\omegaOrd$ (in the case of
    a finite word), less than $2 \cdot \omegaOrd$ (in the case of an infinite
    word thanks to \cref{small-ordinal-invariants:lem}), or less than $3 \cdot
    \omegaOrd$ (in the case of a bi-infinite word, thanks to
    \cref{from-bi-to-single:lem}). In any case,
    we have the bound $\oWidth{L_i} < 3 \cdot \omegaOrd$.

>>>>

We are now ready to conclude the proof of \cref{dw-closed-wqo-charac:thm}.

\begin{proofof}{dw-closed-wqo-charac:thm}
  It is clear that the set of finite \kl{infixes} of a finite, infinite or
  bi-infinite \kl{ultimately uniformly recurrent} word is \kl{well-quasi-ordered}
  for the \kl{infix relation} thanks to \cref{ultimately-uniformly-recurrent:lem}.

@@

Finally, we comment on the \kl{ordinal invariants} of the set of finite
\kl{infixes} of an \kl{ultimately uniformly recurrent} infinite word, in order
to derive the bounds of \cref{small-ordinal-invariants:thm}.

\begin{lemma}
    \label{small-ordinal-invariants:lem}
    \proofref{small-ordinal-invariants:lem}
    Let $w \in \Sigma^\Nat$ be an \kl{ultimately uniformly recurrent} word.
    Then, the set of finite infixes of $w$ has \kl{ordinal width}
    less than $2 \cdot \omegaOrd$.
    Furthermore, this bound is tight. 
\end{lemma}

\begin{lemma}
    \label{from-bi-to-single:lem}
    \proofref{from-bi-to-single:lem}
    Let $w \in \Sigma^\Rel$ be a bi-infinite word. Then, the set of finite
    infixes of $w$ is \kl{well-quasi-ordered} for the \kl{infix relation} if
    and only if $w_+$ and $w_-$ are two \kl{ultimately uniformly recurrent}
    words. In this case, the \kl{ordinal width}
    of the set of finite infixes of $w$ is less than $3 \cdot \omegaOrd$,
    and this bound is tight.
\end{lemma}

Ordinal invariants in hand, we proceed with the proof of \cref{small-ordinal-invariants:thm}.
\begin{proofof}{small-ordinal-invariants:thm}
    It is always true that the \kl{ordinal height} of a language over a finite
    alphabet is at most $\omegaOrd$. Let us now consider a
    \kl{well-quasi-ordered} language $L$ that is \kl{downwards closed} for the
    \kl{infix relation}. Applying 
    \cref{dw-closed-wqo-charac:thm}, we can write $L = \bigcup_{i = 1}^n L_i$ where
    each $L_i$ is the set of finite \kl{infixes} of a finite, infinite or
    bi-infinite \kl{ultimately uniformly recurrent} word $w_i$.
   We can then
    directly conclude that $\oWidth{L_i}$ less than $\omegaOrd$ (in the case of
    a finite word), less than $2 \cdot \omegaOrd$ (in the case of an infinite
    word thanks to \cref{small-ordinal-invariants:lem}), or less than $3 \cdot
    \omegaOrd$ (in the case of a bi-infinite word, thanks to
    \cref{from-bi-to-single:lem}). In any case,
    we have the bound $\oWidth{L_i} < 3 \cdot \omegaOrd$.
